\textbf{Efficient Implementation Strategies for Block Ciphers on ARMv8}\\ {\footnotesize Bachelorarbeit}

### 2023-05-19 Paper/chapters/appendix.tex
finish draft

\section{Implementations for SPN}
\subsection{Table-based}

\lstinputlisting[language=c, caption={\texttt{gift\_table.h}}]{../impl/gift/table.h}
\lstinputlisting[language=c, caption={\texttt{gift\_table.c}}]{../impl/gift/table.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sbox.h}}]{../impl/gift/vec_sbox.h}
\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sbox.c}}]{../impl/gift/vec_sbox.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sliced.h}}]{../impl/gift/vec_sliced.h}
\lstinputlisting[language=c, caption={\texttt{gift\_vec\_sliced.c}}]{../impl/gift/vec_sliced.c}

\chapter{Lorem dolor}

>>>>

\section{GIFT}
\subsection{Table-based}

\lstinputlisting[language=c, caption={\texttt{gift/table.h}}]{../impl/gift/table.h}
\lstinputlisting[language=c, caption={\texttt{gift/table.c}}]{../impl/gift/table.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift/vec\_sbox.h}}]{../impl/gift/vec_sbox.h}
\lstinputlisting[language=c, caption={\texttt{gift/vec\_sbox.c}}]{../impl/gift/vec_sbox.c}

@@

\lstinputlisting[language=c, caption={\texttt{gift/vec\_sliced.h}}]{../impl/gift/vec_sliced.h}
\lstinputlisting[language=c, caption={\texttt{gift/vec\_sliced.c}}]{../impl/gift/vec_sliced.c}

\subsection{Optimized non-SIMD}

\lstinputlisting[language=c, caption={\texttt{camellia/spec\_opt.h}}]{../impl/camellia/spec_opt.h}
\lstinputlisting[language=c, caption={\texttt{camellia/spec\_opt.c}}]{../impl/camellia/spec_opt.c}

\subsection{Byteslicing}

\lstinputlisting[language=c, caption={\texttt{camellia/bytesliced.h}}]{../impl/camellia/bytesliced.h}
\lstinputlisting[language=c, caption={\texttt{camellia/bytesliced.c}}]{../impl/camellia/bytesliced.c}

### 2023-05-19 Paper/chapters/implementation-strategies.tex
finish draft

Bitsliced implementation techniques were first introduced to improve the performance
of DES in 1997 and work by viewing a processor with $n$-bit registers as a
machine capable of executing $n$ bitwise operations at
once\cite{desslicing:1997}. Bitslicing offers a performance advantage by
splitting up $n$ bits into $m$ slices to achieve a more efficient
representation which can exploit this bitwise parallelism. The structure of
\texttt{GIFT} naturally offers possibilities for bitslicing. We split the
cipher state bits $b_{63}b_{62}\dots b_0$ into four slices $S_i,
i\in\{0,1,2,3\}$ such that the $i$-th slice contains all $i$-th bits of the
individual S-boxes. This is equivalent to transposing the bit matrix.

@@

$n=m=1$ would be a valid choice which yields eight messages divided into four
registers. We choose $n=m=2$ so we can directly utilize the algorithm for bit
packing presented by the original GIFT authors, although it is simple to adapt
this algorithm to only four registers and eight messages by adjusting the
\texttt{SWAPMOVE} shift and mask values.

@@

\subsection{Platform-independent techniques}

@@

with affine transformations $h,f$ and the multiplicative inversion function $g$
in the composite field GF$((2^4)^2)$. Because Galois fields of equal size are
isomorphic, there exist affine isomorphisms $\delta,\delta^{-1}$ between
GF($2^8$) and GF($(2^4)^2$) respectively\cite{satohmorioka:2003}:

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\textwidth]{Figures/camellia_bytesliced_rol.pdf}
    \caption{Camellia bytesliced 1-bit left rotate}
\end{figure}

>>>>

Bitsliced implementation techniques were first introduced to improve the
performance of DES in 1997 and work by viewing a processor with $n$-bit
registers as a machine capable of executing $n$ bitwise operations at
once \cite{desslicing:1997}. Bitslicing offers a performance advantage by
splitting up $n$ bits into $m$ slices to achieve a more efficient
representation which can exploit this bitwise parallelism. The structure of
\texttt{GIFT} naturally offers possibilities for bitslicing. We split the
cipher state bits $b_{63}b_{62}\dots b_0$ into four slices $S_i,
i\in\{0,1,2,3\}$ such that the $i$-th slice contains all $i$-th bits of the
individual S-boxes. This is equivalent to transposing the bit matrix.

@@

$n=m=1$ would be a valid choice which yields eight messages divided into four
registers. We choose $n=m=2$ so we can directly utilize the algorithm for bit
packing presented by the original GIFT authors, although it is simple to adapt
this algorithm to only four registers and eight messages by adjusting the
\texttt{SWAPMOVE} shift and mask values. Encrypting multiple blocks in parallel
is a realistic use case since usually a lot of blocks are encrypted under the
same key.

@@

\subsection{Optimized non-SIMD}

@@

with affine transformations $h,f$ and the multiplicative inversion function $g$
in the composite field GF$((2^4)^2)$. Because Galois fields of equal size are
isomorphic, there exist affine isomorphisms $\delta,\delta^{-1}$ between
GF($2^8$) and GF($(2^4)^2$) respectively \cite{satohmorioka:2003}:

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.8\textwidth]{Figures/camellia_bytesliced_rol.pdf}
    \caption{Camellia bytesliced 1-bit left rotate}
\end{figure}

### 2023-05-19 Paper/chapters/introduction.tex
rename files, add implementation sample code
finish draft

\texttt{GIFT}\cite{gift:2017}, first presented in the \textit{CHES 2017}
cryptographic hardware and embedded systems conference, is a lightweight block
cipher based on a previous design called \texttt{PRESENT}, developed in 2007. Its
goal is to offer maximum security while being extremely light on resources.
Modern battery-powered devices like RFID tags or low-latency operations like
on-the-fly disc encryption present strong hardware and power constraints. GIFT
aims to be a simple, low-energy cipher suited for these kinds of applications.

@@

Camellia\cite{camellia:2001} is a block cipher jointly developed by NTT and
Mitsubishi Electric Corporation and first published in 2001. Following AES
specifications, it is able to encrypt 128-bit blocks using either 128-, 196- or
256-bit keys and claims to possess similar performance and security levels as
the AES finalists.

@@

With small devices and embedded processors becoming ever more ubiquitous and
essential in areas like consumer electronics or industrial and IoT
applications, the need for low-power, high-performance microprocessors has
increased steadily. With more than 250 billion chips shipped, semiconductors
designed by ARM power 95\% of mobile devices and have found a great many
applications due to their high performance and low power
consumption\cite{armcompany}. The ODROID-N2+\cite{odroidn2} development board
we are using is based on the big.LITTLE architecture and is powered by a
quad-core ARM Cortex-A73 processor and a weaker dual-core ARM Cortex-A53 for
power efficiency. Both these processors are part of the eight generation of ARM
designs known as ARMv8\cite{armv8:2013}.

@@

\begin{table}[h!]
    \centering
    \footnotesize
    \caption{Common NEON intrinsics}
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Intrinsic && Summary \\
        \midrule
        \texttt{uint64\_t} & \texttt{vgetq\_lane\_u64(void)} & Extract a single lane \\
        \midrule
        \texttt{void} & \texttt{vsetq\_lane\_u64(uint64\_t)} & Insert a single lane \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vdupq\_n\_u64(uint64\_t)} & Initialize all lanes to same value \\
        \midrule
        \texttt{void} & \texttt{vst1q\_u64(uint64\_t*, uint64x2\_t)} & Store from register to memory \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vld1q\_u64(uint64\_t*, uint64x2\_t)} & Load from memory to register \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{veorq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise XOR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vandq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise AND \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vorrq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise OR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vmvnq\_u8(uint8x16\_t)} & bitwise NOT \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vqtbl2q\_u8(uint8x16\_t, uint8x16\_t)} & permutation (\texttt{TBL}) \\
        \bottomrule
    \end{tabularx}
\end{table}

>>>>

\texttt{GIFT} \cite{gift:2017}, first presented in the \textit{CHES 2017}
cryptographic hardware and embedded systems conference, is a lightweight block
cipher based on a previous design called \texttt{PRESENT}, developed in 2007. Its
goal is to offer maximum security while being extremely light on resources.
Modern battery-powered devices like RFID tags or low-latency operations like
on-the-fly disc encryption present strong hardware and power constraints. GIFT
aims to be a simple, low-energy cipher suited for these kinds of applications.

@@

Camellia \cite{camellia:2001} is a block cipher jointly developed by NTT and
Mitsubishi Electric Corporation and first published in 2001. Following AES
specifications, it is able to encrypt 128-bit blocks using either 128-, 196- or
256-bit keys and claims to possess similar performance and security levels as
the AES finalists.

@@

With small devices and embedded processors becoming ever more ubiquitous and
essential in areas like consumer electronics or industrial and IoT
applications, the need for low-power, high-performance microprocessors has
increased steadily. With more than 250 billion chips shipped, semiconductors
designed by ARM power 95\% of mobile devices and have found a great many
applications due to their high performance and low power consumption
\cite{armcompany}. The ODROID-N2+ \cite{odroidn2} development board we are
using for testing is powered by an Amlogic S922X processor with four ARM
Cortex-A73 and two weaker ARM Cortex-A53 cores based on the big.LITTLE
architecture combining slower, more power efficient cores with faster, more
power hungry ones. Both these cores are part of the eight generation of ARM
designs known as ARMv8 \cite{armv8:2013} released in 2011 and succeeding the
older 32-bit ARMv7.

@@

\begin{table}[h!]
    \centering
    \footnotesize
    \caption{Common NEON intrinsics}
    \begin{tabularx}{\textwidth}{llX}
        \toprule
        Intrinsic && Summary \\
        \midrule
        \texttt{uint64\_t} & \texttt{vgetq\_lane\_u64(void)} & Extract a single lane \\
        \midrule
        \texttt{void} & \texttt{vsetq\_lane\_u64(uint64\_t)} & Insert a single lane \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vdupq\_n\_u64(uint64\_t)} & Initialize all lanes to same value \\
        \midrule
        \texttt{void} & \texttt{vst1q\_u64(uint64\_t*, uint64x2\_t)} & Store from register to memory \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vld1q\_u64(uint64\_t*, uint64x2\_t)} & Load from memory to register \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{veorq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise XOR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vandq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise AND \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vorrq\_u8(uint8x16\_t, uint8x16\_t)} & bitwise OR \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vmvnq\_u8(uint8x16\_t)} & bitwise NOT \\
        \midrule
        \texttt{uint8x16\_t} & \texttt{vqtbl2q\_u8(uint8x16\_t, uint8x16\_t)} & permutation (\texttt{TBL}) \\
        \midrule
        \texttt{uint64x2\_t} & \texttt{vextq\_u64(uint64x2\_t, uint64x2\_t, int)} & extract from pair of vectors \\
        \bottomrule
    \end{tabularx}
\end{table}
